\bf What is the Best Way to Display Two Time Series to Read Association?

### 2016-01-20 paper/paper.Rnw
lots more to do in section 2, but stopping for now

The lineup protocol equips exploratory visual methods with a rigorous procedure that enables statistical inference. This is important because visualization remains essential for data analysis because human eyes can detect patterns which would not be detected mathematically. The lineup protocol provides the calibration of the human eyes to prevent us from erroneously jumping to conclusions about magic patterns. Earlier than the protocol proscription, we can find preliminary ideas along these lines, for example \citet{Daniel} recommends looking at plots of data from a null distribution for exploring two factor design experimental data.  His paper provides 40 pages of plots from the null! He encourages data analysts to understand the patterns that might be seen in samples data drawn from populations without inherent structure. (\citet{Buja} calls this the Rorschach protocol.)

@@

With the general lack of inference present in exploratory data analysis, how do we attempt to control error in hypothesis testing? In order to use visualisation effectively, we will get the most benefit if the type of graphic display chosen is best suited to the task at hand, be it EDA, MD, or even the presentation of results.
We then need to find the statistical power of graphics, the ability of a graphical display to convey information on the structure of the data within. This can be found by using the lineup protocol developed in 2009 by \citet{Buja}, which is easily implemented with the nullabor R package.
They created a statistically rigorous framework with properties explored primarily in \citet{Majumder:2014} and further in \citet{Hofmann:2014} that allows us to conduct these hypothesis tests with visual inference, thus to some degree allowing the conjoining of EDA with classical inferential statistics.
The lineup protocol is inspired in part by the police lineup \citep{Wickham}. We place the plot of the 'true' data generated with some underlying structure (The criminal) in amongst plots of data that were generated from a null distribution (The innocent people). If the real data plot is selected by an uninvolved observer as being most different from the other plots there is evidence that the structure of that data has led to a significant difference in the plot (That the criminal is sufficiently different from the innocents). 
This constitutes a rejection of the null hypothesis. If one of the null plots is chosen instead then either the plot design did not have the sufficient power to display the true relationship (A Type 2 error in EDA), or the null plot exhibited a strong relationship that was generated randomly and an analyst that saw this plot and decided that the data had some structure would’ve committed a Type 1 error. 
$p_{i}$, the probability that plot $i$ is chosen in the lineup depends not only on the plot design $d$, but also the signal strength, $q_{i}$, of that plot and of every other competing plot in the lineup. It can be defined as some unknown function $f_{i,d} (q_{1},\dots,q_{20})$ \citep{Hofmann:2014}. 
If the true plot is detected, it indicates that the plot design could convey the desired information about the underlying structure and that the true plot has a greater signal strength than those generated under the null distribution. The plot design that has a human observer selecting the true plot the most often will thus minimise both Type 1 error and Type 2 error in EDA hypothesis testing.
However, there is the possibility that the true plot was selected by chance, that we have committed a Type 1 error in the lineup test. For a lineup of $m$ plots, the probability of selecting any plot when they are all generated by the null distribution is $1/m$, setting the Type 1 error rate, $\alpha$, of a lineup hypothesis test to equal $1/m$. 
To control our error rate, it is initially obvious that we can simply change $m$, with an increase in null lineups having an inverse reduction in $\alpha$. We recruit human observers to judge the lineups, but this can quickly lead to a large cognitive burden to sort through more and more null plots. A much more powerful option is to have multiple different viewers of each lineup, with $K$ observers; the probability that at a particular plot was selected at least $x$ times under the null hypothesis is binomially distributed \citep{Buja, Hofmann:2014} with:

@@

@@

>>>>

The lineup protocol equips exploratory visual methods with a rigorous procedure that enables statistical inference. This is important because visualization remains essential for data analysis because human eyes can detect patterns which would not be detected mathematically. The lineup protocol provides the calibration of the human eyes to prevent us from erroneously jumping to conclusions about magic patterns. Prior to the protocol description, we can find preliminary ideas along these lines, for example \citet{Daniel} recommends looking at plots of data from a null distribution for exploring two factor design experimental data.  His paper provides 40 pages of plots from the null! He encourages data analysts to understand the patterns that might be seen in samples data drawn from populations without inherent structure. (\citet{Buja} has named this the Rorschach protocol.)

@@

%With the general lack of inference present in exploratory data analysis, how do we attempt to control error in hypothesis testing? In order to use visualisation effectively, we will get the most benefit if the type of graphic display chosen is best suited to the task at hand, be it EDA, MD, or even the presentation of results.
%We then need to find the statistical power of graphics, the ability of a graphical display to convey information on the structure of the data within. This can be found by using the lineup protocol developed in 2009 by \citet{Buja}, which is easily implemented with the nullabor R package.
%They created a statistically rigorous framework with properties explored primarily in \citet{Majumder:2014} and further in \citet{Hofmann:2014} that allows us to conduct these hypothesis tests with visual inference, thus to some degree allowing the conjoining of EDA with classical inferential statistics.
%The lineup protocol is inspired in part by the police lineup \citep{Wickham}. We place the plot of the 'true' data generated with some underlying structure (The criminal) in amongst plots of data that were generated from a null distribution (The innocent people). If the real data plot is selected by an uninvolved observer as being most different from the other plots there is evidence that the structure of that data has led to a significant difference in the plot (That the criminal is sufficiently different from the innocents). 
%This constitutes a rejection of the null hypothesis. If one of the null plots is chosen instead then either the plot design did not have the sufficient power to display the true relationship (A Type 2 error in EDA), or the null plot exhibited a strong relationship that was generated randomly and an analyst that saw this plot and decided that the data had some structure would’ve committed a Type 1 error.

% First the p-value calculation for a lineup, multiple observers

The lineup protocol specifies that the data plot be placed in a random location in a field of $(m-1)$ null plots. Let $p_{i}$ be the probability that plot $i$ is chosen from the lineup. It depends on the plot design, $d$, and the signal strength, $q_{i}$, of the structure displayed in plot relative to the other plots in the lineup. This can be defined as a function $f_{i,d} (q_{1},\dots,q_{20})$\citep{Hofmann:2014}. If the data plot is detected, it indicates that the plot design could convey the desired information about the underlying structure and that the data plot has a higher signal strength than the null plots. The plot design that has a human observer selecting the data plot the most often will thus minimise both Type I error and Type II error in the hypothesis testing.

However, there is the possibility that the true plot was selected by chance, that we have committed a Type I error in the lineup test. For a lineup of $m$ plots, the probability of selecting any plot when they are all generated by the null distribution is $1/m$, setting the Type I error rate, $\alpha$, of a lineup hypothesis test to equal $1/m$. 
To control our error rate, we could increase change $m$, this could produce a large cognitive burden on one observer, having to sort through many plots. An alternative approach is to recruit more observers, $K$, to evaluate a lineups, then the probability that plot $i$ is selected by $x$ observers under the null hypothesis follows a binomial distribution \citep{Majumder:2013}:

@@

% Multiple picks

@@

% Then power calculation for comparing designs
